
---
title: "W271 Lab 2"
author: "Akin Akinlabi, Eduardo Gonzalez, Karanveer Lamba, Rui Sun, Victor Ndayambaje"
output: pdf_document
date: '`r Sys.Date()`'
---

```{r setup, include=FALSE}
# Loading Libraries
knitr::opts_chunk$set(echo = TRUE)
library(stargazer)
library(ggplot2)
library(dplyr)
library(tidyr)
library(httr)
library(jsonlite)
library(fredr)
library(lubridate)
library(forecast)
```

## Task 1: Data Retrieval and Initial EDA (Quarterly Frequency)
```{r Data Loading}

# I've used the fredr package in R to pull data - https://cran.r-project.org/web/packages/fredr/vignettes/fredr.html

# Set FRED API Key object
fredr_set_key("c58e9c0793ba6ec253e8588464dd1eca")

# Set date range
start_date <- as.Date("1980-01-01")
end_date <- Sys.Date()

# Download Federal Funds Rate (FEDFUNDS)
fed_funds <- fredr(series_id = "FEDFUNDS", observation_start = start_date, frequency = "q") %>%
  select(date, fed_rate = value)

# Download PCE Price Index (PCEPI) for Inflation
pcepi <- fredr(series_id = "PCEPI", observation_start = start_date, frequency = "q") %>%
  select(date, pce_index = value)

# Download Real GDP (GDPC1)
real_gdp <- fredr(series_id = "GDPC1", observation_start = start_date, frequency = "q") %>%
  select(date, real_gdp = value)

# Download Potential GDP (GDPPOT)
potential_gdp <- fredr(series_id = "GDPPOT", observation_start = start_date, frequency = "q") %>%
  select(date, potential_gdp = value)

```

``` {r create inflation series}

pcepi <- pcepi %>%
  arrange(date) %>%
  mutate(inflation = round((pce_index / lag(pce_index) - 1) * 100, 2))

```

``` {r create output gap series}

output_gap <- real_gdp %>%
  inner_join(potential_gdp, by = "date") %>%
  mutate(output_gap = ((real_gdp - potential_gdp) / potential_gdp) * 100)

```

``` {r merge data}

df <- fed_funds %>%
  inner_join(pcepi, by = "date") %>%
  inner_join(output_gap, by = "date")
  
```

``` {r plot series}

# Federal Funds Rate over time
ggplot(df, aes(x = date, y = fed_rate)) +
  geom_line(color = "blue") +
  labs(title = "Federal Funds Rate Over Time", x = "Year", y = "Fed Funds Rate (%)") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +  # Set 2-year gaps on X-axis
  theme_minimal()

# Inflation over time
ggplot(df, aes(x = date, y = inflation)) +
  geom_line(color = "red") +
  labs(title = "Inflation Rate Over Time", x = "Year", y = "Inflation (%)") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +  # Set 2-year gaps on X-axis
  theme_minimal()

# Output Gap over time
ggplot(df, aes(x = date, y = output_gap)) +
  geom_line(color = "green") +
  labs(title = "Output Gap Over Time", x = "Year", y = "Output Gap (%)") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +  # Set 2-year gaps on X-axis
  theme_minimal()
```
``` {r summary statistics}

summary(df)

```

### Plot Interpretation

#### Federal Funds Rate (Top Panel - Blue Line)

- 1980s: Interest rates were at historically high levels (above 15%) due to the Federal Reserve’s aggressive tightening to combat inflation.

- 2000s: The rate remained relatively stable before being cut sharply after the 2008 Financial Crisis, when the Fed reduced rates to near-zero to stimulate the economy.

- 2020: Another near-zero rate policy followed during the COVID-19 pandemic to counter economic downturns.

- 2022-Present: A rapid rate hike is visible, reflecting the Fed’s response to post-pandemic inflation.

#### Inflation Rate (Middle Panel - Red Line)

- 1980s: Inflation started high but steadily declined after the Volcker-led Fed raised interest rates aggressively.

- 2008 Financial Crisis: There was a brief period of deflation (inflation dropping below zero).

- 2020s: A sharp spike in inflation post-COVID due to supply chain disruptions, stimulus spending, and demand recovery.

#### Output Gap (Bottom Panel - Green Line)

- 1980s-1990s: Large negative output gaps were seen during recessions, indicating economic contractions.

- 2008-2009 Financial Crisis: A sharp drop in the output gap occurred, reflecting a deep recession.

- 2020 (COVID-19 Recession): The steepest drop in output gap, followed by a strong recovery.

## Task 2: Preprocessing 

All datasets are set to be quarterly using as.yearqtr(date), and are aggregated from the monthly values to quarterly with the aggregate function.

```{r frequency alignment}
library(zoo)
df$quarter <- as.yearqtr(df$date)  # Create a quarter column
df_quarterly <- aggregate(cbind(fed_rate, pce_index, inflation, real_gdp, potential_gdp, output_gap) ~ quarter, data = df, FUN = mean)
summary(df_quarterly)
```

From task 1, the three plots show extreme spikes, indicating outliers. Winsorizing is a method to cap extreme outliers. Instead of removing outliers, in this case to retain all data points but to modify extreme values and their influence, any data point above the 99th percentile are capped at the 99th percentile value, and any data point below the 1st percentile are capped at the 1st percentile value.

```{r outlier detection and treatment}
# Set percentile thresholds for Winsorizing (1st and 99th percentile)
lower_percentile <- 0.01
upper_percentile <- 0.99

# Function to Winsorize a variable
winsorize <- function(x, lower_percentile, upper_percentile) {
  lower_cutoff <- quantile(x, lower_percentile)
  upper_cutoff <- quantile(x, upper_percentile)
  
  # Cap values at the 1st and 99th percentiles
  x[x < lower_cutoff] <- lower_cutoff
  x[x > upper_cutoff] <- upper_cutoff
  
  return(x)
}

# Apply Winsorizing to all variables
df_quarterly$fed_rate <- winsorize(df_quarterly$fed_rate, lower_percentile, upper_percentile)
df_quarterly$pce_index <- winsorize(df_quarterly$pce_index, lower_percentile, upper_percentile)
df_quarterly$inflation <- winsorize(df_quarterly$inflation, lower_percentile, upper_percentile)
df_quarterly$real_gdp <- winsorize(df_quarterly$real_gdp, lower_percentile, upper_percentile)
df_quarterly$potential_gdp <- winsorize(df_quarterly$potential_gdp, lower_percentile, upper_percentile)
df_quarterly$output_gap <- winsorize(df_quarterly$output_gap, lower_percentile, upper_percentile)
```

The ggplot2 package is used to visualize the real GDP over time, allowing for the identification of any potential seasonal variation. To further assess seasonality, the nsdiffs function from the forecast package is used to test for the need for seasonal differencing. STL decomposition is applied using the stl() function, which separates the time series into its trend, seasonal, and residual components. The seasonal component is extracted to assess potential seasonal effects. Based on the seasonal plot and a seasonality test result of zero (indicating no significant seasonality), combined with the STL decomposition showing a general upward trajectory, repeating seasonal patterns, and a large spike in the residual component (which cannot be explained by the seasonality), it is concluded that there is no significant seasonal effect on the dataset. Therefore, seasonality adjustment is not necessary.

```{r seasonal adjustment}
# Seasonal plot of real_gdp
library(ggplot2)
ggplot(df_quarterly, aes(x = quarter, y = real_gdp)) +
  geom_line() +
  facet_wrap(~ factor(month(quarter)), scales = "free_y") +
  labs(title = "Seasonal Plot of Real GDP", x = "Quarter", y = "Real GDP")

library(forecast)

# Convert the data to a time series object (quarterly data)
ts_data <- ts(df_quarterly$real_gdp, frequency = 4)

# Test for seasonal differencing
seasonality_test <- nsdiffs(ts_data)

# Print the result
print(seasonality_test)

# Apply STL decomposition
decomp <- stl(ts_data, s.window = "periodic")

# Plot the decomposition to visualize the seasonal component
plot(decomp)
```

## Task 3: Stationarity Testing & Potential Structural Breaks

The Augmented Dickey-Fuller tests for unit roots were conducted with three different regression models: none, drift, and trend. The trend model, which included both a constant and a linear trend, had a test statistic of -2.18, also showing no significant evidence of stationarity, with a p-value of 0.05779. The second model (drift) included a constant, and the test statistic of 1.17 showed no significant evidence of stationarity, with a p-value of 0.2194. Finally, the test statistic was 5.44, which is significant at the 1% level, suggesting the series is stationary. This progression from the most general model (trend) to the simplest model (none) supports the conclusion that the series is stationary.

```{r stationarity with structural break}
library(aTSA)
library(urca)

ts_quarterly <- ts(df_quarterly$real_gdp, start = c(1980, 1), frequency = 4)

# ADF test with constant and trend, specifying 4 lags
adf_trend <- ur.df(ts_quarterly, type = "trend", lags = 4)
summary(adf_trend)

# ADF test with constant (drift), specifying 4 lags
adf_drift <- ur.df(ts_quarterly, type = "drift", lags = 4)
summary(adf_drift)

# ADF test with no constant or trend, specifying 4 lags
adf_none <- ur.df(ts_quarterly, type = "none", lags = 4)  # Specify a valid number of lags
summary(adf_none)
```
## Task 4

```{r stationarity with structural break}
# Assuming df_quarterly is your dataframe and it has a column named 'inflation'
df_quarterly <- df_quarterly %>%
  mutate(inflation_gap = inflation - 2)

# Add a second predictor: lagged value of the time series
df_quarterly <- df_quarterly %>%
  mutate(lagged_inflation = lag(inflation, order_by = quarter))
train_set <- df_quarterly %>% filter(quarter < '2022 Q4')
test_set <- df_quarterly %>% filter(quarter >= '2022 Q4')

# Remove NA values
train_set <- na.omit(train_set)
test_set <- na.omit(test_set)

# Fit the ols with lag model
ols_model <- glm(fed_rate ~ lagged_inflation + inflation_gap + output_gap , data = train_set)

# Fit the ols without lag model
ols_model <- glm(fed_rate ~ inflation_gap + output_gap , data = train_set)

# Predict on the test set
predictions <- predict(ols_model, newdata = test_set)

# Calculate the residuals
residuals <- residuals(ols_model)

# Summarize the model
summary(ols_model)
# Predict on the training set
train_set$Predicted <- predict(ols_model, newdata = train_set)

# Predict on the test set
test_set$Predicted <- predict(ols_model, newdata = test_set)

# Calculate the residuals
residuals <- residuals(ols_model)

# Calculate the RMSE
rmse <- sqrt(mean((train_set$fed_rate - train_set$Predicted)^2))
print(paste("In Sample: Root Mean Squared Error (RMSE):", rmse))
rmse <- sqrt(mean((test_set$fed_rate - test_set$Predicted)^2))
print(paste("Out of Sample: Root Mean Squared Error (RMSE):", rmse))



# Plot the time series and the fitted linear model for the training set
ggplot(train_set, aes(x = quarter)) +
  geom_line(aes(y = fed_rate), color = "black") +
  geom_line(aes(y = Predicted), color = "blue") +
  labs(title = "Linear Model Fit to Time Series Data (Training Set)",
       x = "Time",
       y = "Fed Rate")

# Plot the time series and the fitted linear model for the training set
ggplot(test_set, aes(x = quarter)) +
  geom_line(aes(y = fed_rate), color = "black") +
  geom_line(aes(y = Predicted), color = "blue") +
  labs(title = "Linear Model Fit to Time Series Data (Training Set)",
       x = "Time",
       y = "Fed Rate")

# Calculate the residuals
residuals <- residuals(ols_model)

par(mfrow=c(2,2))

# Plot the residuals using an ACF plot
acf(residuals, main = "ACF of Residuals")
pacf(residuals, main = "PACF of Residuals")
# Plot the histogram of residuals with a normal distribution curve
hist(residuals, breaks = 10, probability = TRUE, main = "Histogram of Residuals with Normal Curve", xlab = "Residuals")
curve(dnorm(x, mean = mean(residuals), sd = sd(residuals)), col = "blue", lwd = 2, add = TRUE, yaxt = "n")
```
According to the https://www.federalreserve.gov/economy-at-a-glance-inflation-pce.htm target inflation is 2%

1. At a parameter alpha0 = 1.89988 and p_value < .05, interest moves more than one for one with inflation. 
Although it does not agree with Taylor's rule, it is worth noting that this model is an oversimplifcation of complex/unpredictable economic variables such as COVID-19
2. In sample RMSE yields 3.27043042498184. 3 percent in interest rates is a not a good sign for the model.
The plot shows that COVID-19 years had a huge impact in the residuals. It may be worth treating COVID-19 data to better the model. Also worth noting is the ACF plot indicates some degree in autocorrelation of the residuals.
3. It is worth mentioning that this is a good prediction of the model as the RSME = 1.37524946670628 is lower in the test set. Evaluating the plot shows the residuals are lessened in comparison to the test data

## Task 5

```{r stationarity with structural break}
library(urca)

# Engle-Granger Cointegration Test
long_run_model <- lm(fed_rate ~ inflation_gap + output_gap, data = train_set)
summary(long_run_model)

# Extract residuals from long-run model
train_set <- train_set %>%
  mutate(residuals = residuals(long_run_model))

# ADF test on residuals (Engle-Granger test)
adf_test_residuals <- ur.df(train_set$residuals, type = "none", lags = 4)
summary(adf_test_residuals)

# Error Correction Model (ECM)
# Compute first differences and lagged residuals
train_set <- train_set %>%
  mutate(
    d_fed_rate = c(NA, diff(fed_rate, lag = 1)),  # Prepend NA to match length
    d_inflation_gap = c(NA, diff(inflation_gap, lag = 1)),
    d_output_gap = c(NA, diff(output_gap, lag = 1)),
    lagged_residuals = lag(residuals, n = 1)
  ) %>%
  na.omit()  # Remove first-row NA values

# Fit the ECM model
ecm_model <- lm(d_fed_rate ~ d_inflation_gap + d_output_gap + lagged_residuals, data = train_set)
summary(ecm_model)

test_set <- test_set %>%
  mutate(
    residuals = predict(long_run_model, newdata = test_set) - test_set$fed_rate
  )

# ECM Forecast Performance on Test Set
test_set <- test_set %>%
  mutate(
    d_fed_rate = c(NA, diff(fed_rate, lag = 1)),
    d_inflation_gap = c(NA, diff(inflation_gap, lag = 1)),
    d_output_gap = c(NA, diff(output_gap, lag = 1)),
    lagged_residuals = lag(residuals, n = 1)
  ) %>%
  na.omit()

# Predict using ECM model
test_set$ECM_Predicted <- predict(ecm_model, newdata = test_set)

# Compute RMSE for ECM model
rmse_ecm_train <- sqrt(mean(residuals(ecm_model)^2))
rmse_ecm_test <- sqrt(mean((test_set$fed_rate - test_set$ECM_Predicted)^2))

print(paste("ECM In-Sample RMSE:", rmse_ecm_train))
print(paste("ECM Out-of-Sample RMSE:", rmse_ecm_test))

# Plot actual vs. predicted (ECM)
ggplot(test_set, aes(x = quarter)) +
  geom_line(aes(y = fed_rate), color = "black") +
  geom_line(aes(y = ECM_Predicted), color = "blue") +
  labs(title = "ECM Model Predictions vs. Actual Fed Funds Rate",
       x = "Quarter",
       y = "Fed Funds Rate (%)")
```
The results shows that the basic Taylor Rule has some explanatory power but lacks a strong fit. The inflation gap is highly significant, with a coefficient of 4.38, meaning that for every 1% increase in the inflation gap, the Fed Funds Rate is expected to rise by about 4.38 percentage points. However, the output gap is statistically insignificant, p = 0.867, suggesting it has little influence on interest rate decisions in this model. The R² of 0.27 indicates that the model explains only 27% of the variation in interest rates, implying that other factors influence Fed policy.
The Augmented Dickey-Fuller test on the residuals suggests a weak rejection of the unit root hypothesis, test statistic = -2.14, above the 5% critical value of -1.95. This indicates some level of co-integration between the Fed Funds Rate, inflation gap, and output gap, but it is not particularly strong.
The Error Correction Model results show that short-term changes in the output gap significantly impact interest rate changes, p < 0.001, whereas short-term changes in the inflation gap do not, p = 0.83. The lagged residuals coefficient, -0.05199, p < 0.001, confirms that there is an adjustment process towards the long-run equilibrium, but the speed of correction is slow. The ECM in-sample RMSE (0.55) suggests a good fit within the training data, but the out-of-sample RMSE (5.11) is much higher, indicating poor forecasting performance.
The plot confirms this issue, as the ECM predictions (blue line) are consistently lower than the actual Fed Funds Rate (black line), meaning the model significantly under predicts interest rates. This could be due to the poor explanatory power of the output gap, weak co-integration, or changes in Fed policy not captured by the Taylor Rule framework.

## Task 6

During this task, we aim to:

- Extract residuals from the OLS-based Taylor Rule model.
- Fit an ARIMA model to these residuals.
- Generate forecasts using the combined Taylor Rule + ARIMA approach.
- Compare forecast accuracy using RMSE.

### Extract Residuals from the Taylor Rule Model

```{r}
# Extract residuals from the Taylor Rule model
taylor_residuals <- residuals(ols_model)  

# Convert residuals into a time series object
ts_residuals <- ts(taylor_residuals, start = c(1980,1), frequency = 4)

# Plot residuals to check for patterns
autoplot(ts_residuals) +
  ggtitle("Residuals of the Taylor Rule Model") +
  xlab("Year") +
  ylab("Residuals")
```
The residuals plot shows the difference between the actual and predicted Fed Funds Rate from the Taylor Rule regression. There are periods of high residual volatility, particularly in the 1980s and 2008, indicating times when the model struggled to capture interest rate movements. The downward trend in recent years suggests potential structural changes in monetary policy.

### Identify the Best ARIMA Model

We check for residual autocorrelation and identify the best ARIMA model.
```{r}
# Automatically select best ARIMA model
arima_residuals_model <- auto.arima(ts_residuals)

# Display ARIMA model summary
summary(arima_residuals_model)

# Check ACF and PACF plots
par(mfrow=c(1,2))
acf(ts_residuals, main="ACF of Residuals")
pacf(ts_residuals, main="PACF of Residuals")

```
The Autocorrelation Function (ACF) shows significant persistence in residuals, indicating that past errors influence future errors. The Partial Autocorrelation Function (PACF) suggests an ARIMA model with an MA(1) and seasonal AR(2) structure is appropriate. This confirms that the Taylor Rule residuals exhibit serial correlation and require correction.

### Forecast Residuals

We forecast residuals for the next 8 quarters (2023Q1–2024Q4).

```{r}
# Forecast residuals for 8 future quarters (2023Q1–2024Q4)
residual_forecast <- predict(arima_residuals_model, n.ahead = 8)

# Convert forecast to a data frame
residual_forecast_df <- data.frame(
  quarter = seq(max(test_set$quarter) + 0.25, by = 0.25, length.out = 8),
  residual_forecast = residual_forecast$pred
)

# Plot forecasted residuals
ggplot(residual_forecast_df, aes(x = quarter, y = residual_forecast)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Forecasted Residuals from ARIMA Model",
       x = "Year", y = "Residuals") +
  theme_minimal()


```

The selected ARIMA(0,1,1)(2,0,0)[4] model includes a first-order moving average component (MA(1)) and two seasonal autoregressive terms (SAR(2)). The residual forecast plot shows fluctuations but does not exhibit a clear trend. The model's AIC (640.9) and BIC (653.4) suggest a reasonable fit but not a perfect one.

### Combine ARIMA and Taylor Rule Predictions

We use test_set from Task 4 to evaluate forecast accuracy.

```{r}
# Forecast Taylor Rule model for test period
test_set$predicted_fed_rate_taylor <- predict(ols_model, newdata = test_set)

# Ensure ARIMA forecast matches test period length
residual_forecast_values <- residual_forecast$pred[1:nrow(test_set)]  # Extract the right number of values

# Add ARIMA residual forecast to Taylor Rule predictions
test_set$predicted_fed_rate_combined <- test_set$predicted_fed_rate_taylor + residual_forecast_values

# Plot the comparison
ggplot(test_set, aes(x = quarter)) +
  geom_line(aes(y = fed_rate, color = "Actual Fed Rate"), linewidth = 1) +  # Change `size` to `linewidth`
  geom_line(aes(y = predicted_fed_rate_taylor, color = "Taylor Rule Prediction"), linetype = "dashed", linewidth = 1) +
  geom_line(aes(y = predicted_fed_rate_combined, color = "Taylor + ARIMA"), linetype = "dotted", linewidth = 1) +
  labs(title = "Actual vs. Predicted Fed Funds Rate",
       x = "Year", y = "Interest Rate (%)") +
  scale_color_manual(name = "Legend",
                     values = c("Actual Fed Rate" = "black",
                                "Taylor Rule Prediction" = "blue",
                                "Taylor + ARIMA" = "red")) +
  theme_minimal()

```
The Taylor Rule model alone (blue dashed line) closely follows the actual Fed Funds Rate (black line), though it slightly underpredicts in some quarters. The Taylor + ARIMA model (red dotted line) significantly underestimates rates, even producing negative interest rate forecasts, which are unrealistic. This suggests that ARIMA residual correction worsened forecast accuracy by introducing excessive downward adjustments.

### Evaluate Forecast Accuracy

To determine whether the Taylor Rule + ARIMA approach improves forecasting, we compare Root Mean Squared Error (RMSE):

```{r}
# Compute RMSE for Taylor Rule alone
rmse_taylor <- sqrt(mean((test_set$fed_rate - test_set$predicted_fed_rate_taylor)^2, na.rm = TRUE))

# Compute RMSE for Taylor Rule + ARIMA
rmse_combined <- sqrt(mean((test_set$fed_rate - test_set$predicted_fed_rate_combined)^2, na.rm = TRUE))

# Print RMSE results
print(paste("Taylor Rule RMSE:", round(rmse_taylor, 3)))
print(paste("Taylor + ARIMA RMSE:", round(rmse_combined, 3)))

# Check which model performs better
if (rmse_combined < rmse_taylor) {
  print("Taylor Rule + ARIMA improves forecasting accuracy!")
} else {
  print("Taylor Rule alone performs better.")
}

```
The Taylor Rule model alone has a lower RMSE (1.261) compared to Taylor + ARIMA (5.845), meaning that adding ARIMA residual corrections worsened the forecast. Instead of improving accuracy, ARIMA over-adjusted the predictions, leading to a higher error.

**Conclusion**

The Taylor Rule alone remains the better model, as evidenced by its lower RMSE (1.261 vs. 5.845) and closer alignment with actual rates. ARIMA residual correction failed to improve predictions, likely due to model overfitting or structural changes in monetary policy. Future improvements could explore alternative time-series methods such as Vector Autoregression (VAR) or regime-switching models to better capture Federal Reserve decision-making.
