---
title: "Lab 2"
output: pdf_document
date: '`r Sys.Date()`'
---

```{r setup, include=FALSE}
# Loading Libraries
knitr::opts_chunk$set(echo = TRUE)
library(stargazer)
library(ggplot2)
library(dplyr)
library(tidyr)
library(httr)
library(jsonlite)
library(fredr)
library(lubridate)
```

## Task 1: Data Retrieval and Initial EDA (Quarterly Frequency)
```{r Data Loading}

# I've used the fredr package in R to pull data - https://cran.r-project.org/web/packages/fredr/vignettes/fredr.html

# Set FRED API Key object
fredr_set_key("c58e9c0793ba6ec253e8588464dd1eca")

# Set date range
start_date <- as.Date("1980-01-01")
end_date <- Sys.Date()

# Download Federal Funds Rate (FEDFUNDS)
fed_funds <- fredr(series_id = "FEDFUNDS", observation_start = start_date, frequency = "q") %>%
  select(date, fed_rate = value)

# Download PCE Price Index (PCEPI) for Inflation
pcepi <- fredr(series_id = "PCEPI", observation_start = start_date, frequency = "q") %>%
  select(date, pce_index = value)

# Download Real GDP (GDPC1)
real_gdp <- fredr(series_id = "GDPC1", observation_start = start_date, frequency = "q") %>%
  select(date, real_gdp = value)

# Download Potential GDP (GDPPOT)
potential_gdp <- fredr(series_id = "GDPPOT", observation_start = start_date, frequency = "q") %>%
  select(date, potential_gdp = value)

```

``` {r create inflation series}

pcepi <- pcepi %>%
  arrange(date) %>%
  mutate(inflation = round((pce_index / lag(pce_index) - 1) * 100, 2))

```

``` {r create output gap series}

output_gap <- real_gdp %>%
  inner_join(potential_gdp, by = "date") %>%
  mutate(output_gap = ((real_gdp - potential_gdp) / potential_gdp) * 100)

```

``` {r merge data}

df <- fed_funds %>%
  inner_join(pcepi, by = "date") %>%
  inner_join(output_gap, by = "date")
  
```

``` {r plot series}

# Federal Funds Rate over time
ggplot(df, aes(x = date, y = fed_rate)) +
  geom_line(color = "blue") +
  labs(title = "Federal Funds Rate Over Time", x = "Year", y = "Fed Funds Rate (%)") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +  # Set 2-year gaps on X-axis
  theme_minimal()

# Inflation over time
ggplot(df, aes(x = date, y = inflation)) +
  geom_line(color = "red") +
  labs(title = "Inflation Rate Over Time", x = "Year", y = "Inflation (%)") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +  # Set 2-year gaps on X-axis
  theme_minimal()

# Output Gap over time
ggplot(df, aes(x = date, y = output_gap)) +
  geom_line(color = "green") +
  labs(title = "Output Gap Over Time", x = "Year", y = "Output Gap (%)") +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +  # Set 2-year gaps on X-axis
  theme_minimal()
```
``` {r summary statistics}

summary(df)

```

### Plot Interpretation

#### Federal Funds Rate (Top Panel - Blue Line)

- 1980s: Interest rates were at historically high levels (above 15%) due to the Federal Reserve’s aggressive tightening to combat inflation. This was known as the Volcker Shock.

- 2000s: The rate remained relatively stable before being cut sharply after the 2008 Financial Crisis, when the Fed reduced rates to near-zero to stimulate the economy.

- 2020: Another near-zero rate policy followed during the COVID-19 pandemic to counter economic downturns.

- 2022-Present: A rapid rate hike is visible, reflecting the Fed’s response to post-pandemic inflation.

#### Inflation Rate (Middle Panel - Red Line)

- 1980s: Inflation started high but steadily declined after the Volcker-led Fed raised interest rates aggressively.

- 2008 Financial Crisis: There was a brief period of deflation (inflation dropping below zero).

- 2020s: A sharp spike in inflation post-COVID due to supply chain disruptions, stimulus spending, and demand recovery.

#### Output Gap (Bottom Panel - Green Line)

- 1980s-1990s: Large negative output gaps were seen during recessions, indicating economic contractions.

- 2008-2009 Financial Crisis: A sharp drop in the output gap occurred, reflecting a deep recession.

- 2020 (COVID-19 Recession): The steepest drop in output gap, followed by a strong recovery.
